\documentclass[DIV16,twocolumn,10pt]{scrreprt}
\usepackage{paralist}
\usepackage{graphicx}
\usepackage[final]{hcar}

%include polycode.fmt

\begin{document}

\begin{hcarentry}{(GHC Type-checker plugin for kind Nat)}
\report{Christiaan Baaij}
\status{actively developed}
% \participants{(PARTICIPANTS OTHER THAN MYSELF)}% optional
\makeheader

% (WHAT IS IT?)
As of GHC version 7.10, GHC's type checking and inference mechanisms can be enriched by plugins.
This particular plugin enriches GHC's knowledge of arithmetic on the type-level.
Specifically it allows the compiler to reason about \emph{equalities} of types of kind \verb!GHC.TypeLits.Nat!.

GHC's type-checker's knowledge of arithmetic is virtually non-existent: it doesn't know addition is associative and commutative, that multiplication distributes over addition, etc.
In a dependently-typed language, or in Haskell using singleton types, one can provide proofs for these properties and use them to type-check programs that depend on these properties in order to be (type-)correct.
However, most of these properties of arithmetic over natural number are elementary school level knowledge, and it is cumbersome and tiresome to keep on providing and proving them manually.
This type-checker plugin adds the knowledge of these properties to GHC's type-checker.

For example, using this plugin, GHC now knows that:
\begin{verbatim}
(x + 2)^(y + 2)
\end{verbatim}
is equal to:
\begin{verbatim}
4*x*(2 + x)^y + 4*(2 + x)^y + (2 + x)^y*x^2
\end{verbatim}

The way that the plugin works, is that it normalises arithmetic expressions to a normal form that very much resembles \emph{Cantor normal form for ordinals}(\url{http://en.wikipedia.org/wiki/Ordinal_arithmetic#Cantor_normal_form}).
Subsequently, it perform a simple syntactic equality of the two expressions.
Indeed, in the example above, the latter expression is the normal form of the former expression.

% (WHAT IS ITS STATUS? / WHAT HAS HAPPENED SINCE LAST TIME?)
The main test suite for the plugin can be found at: \url{https://github.com/christiaanb/ghc-typelits-natnormalise/blob/master/tests/Tests.hs}.
It demonstrates what kind of \emph{correct} code can be written without type equality annotations, or the use of \verb!unsafeCoerce!.

One important aspect of this plugin is that it only enriches the type checkers knowledge of equalities, but \underline{not} \emph{in}equalities.
That is, it does not allow GHC to solve constraints such as:
\begin{verbatim}
CmpNat (x + 2) (x + 3) ~ 'LT
\end{verbatim}

% (CAN OTHERS GET IT?)
The plugin is available on hackage, for GHC version 7.10 and higher:
\begin{verbatim}
$ cabal update
$ cabal install ghc-typelits-natnormalise
\end{verbatim}

% (WHAT ARE THE IMMEDIATE PLANS?)
Development focus for the plugin is on further testing and improving its testsuite.

\FurtherReading
  \url{http://hackage.haskell.org/package/ghc-typelits-natnormalise} \\
  \url{http://hackage.haskell.org/package/base/docs/GHC-TypeLits.html}
\end{hcarentry}

\end{document}
